---
title: "scDblFinder"
author:
- name: Pierre-Luc Germain
  affiliation: University and ETH Zürich
package: scDblFinder
output:
  BiocStyle::html_document
abstract: |
  An introduction to the scDblFinder method for fast and comprehensive doublet 
  identification in single-cell data.
vignette: |
  %\VignetteIndexEntry{2_scDblFinder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(BiocStyle)
```

# scDblFinder

`scDblFinder` identifies doublets in single-cell RNAseq by creating artificial doublets and looking at their
prevalence (as well as that of any known doublets) in the neighborhood of each cell, along with a few other covariates.
The rough logic is very similar to other methods (e.g. `r Githubpkg("chris-mcginnis-ucsf/DoubletFinder")`, `r Biocpkg("scds")`), 
with a few twists that make it more efficient, more accurate, and provide extra features.

## Installation

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("scDblFinder")

# or, to get that latest developments:
BiocManager::install("plger/scDblFinder")
```

## Usage

The input of `scDblFinder` is an object `sce` of class `r Biocpkg("SingleCellExperiment")` (empty drops having already been removed) containing at least the counts (assay 'counts'). If normalized expression (assay 'logcounts') and PCA (reducedDim 'PCA') are also present, these will be used for the clustering step (unless clusters are given.)

Given an SCE object:

```{r, warning=FALSE}
set.seed(123)
library(scDblFinder)
# we create a dummy dataset; since it's small we set a higher doublet rate
sce <- mockDoubletSCE(dbl.rate=0.1)
# we run scDblFinder (providing the unsually high doublet rate)
sce <- scDblFinder(sce, dbr=0.1)
```

For 10x data, it is usually safe to leave the `dbr` empty, and it will be automatically estimated.
`scDblFinder` will add a number of columns to the colData of `sce` prefixed with 'scDblFinder', the most important of which are:

* `sce$scDblFinder.score` : the final doublet score
* `sce$scDblFinder.class` : the classification (doublet or singlet)

```{r}
table(truth=sce$type, call=sce$scDblFinder.class)
```


### Multiple samples

If you have multiple samples (understood as different cell captures), then it is
preferable to look for doublets separately for each sample (for multiplexed samples with cell hashes, 
this means for each batch). You can do this by simply providing a vector of the sample ids to the 
`samples` parameter of scDblFinder or, if these are stored in a column of `colData`, the name of the 
column. In this case, you might also consider multithreading it using the `BPPARAM` parameter (assuming you've got enough RAM!). 
For example:

```{r, eval=FALSE}
library(BiocParallel)
sce <- scDblFinder(sce, samples="sample_id", BPPARAM=MulticoreParam(3))
table(sce$scDblFinder.class)
```

Note that if you are running multiple samples, clustering will be performed sample-wise. While this is not an issue for doublet identification, it means that the cluster labels (and putative origins of doublets) won't match between samples. If you are interested in these, it is preferable to first cluster (for example using `sce$cluster <- fastcluster(sce)`) and then provide the clusters to `scDblFinder`, which will ensure concordant labels across samples.

<br/><br/>

## Description of the method

Wrapped in the `scDblFinder` function are the following steps:

### Splitting captures

Doublets can only arise within a given sample or capture, and for this reason are better sought independently for each sample, which also speeds up the analysis. If the `samples` argument is given, `scDblFinder` will use it to split the cells into samples/captures, and process each of them in parallel if the `BPPARAM` argument is given. The classifier will be trained globally, but thresholds will be optimized on a per-sample basis. If your samples are multiplexed, i.e. the different samples are mixed in different batches, then the batches should be what you provide to this argument.

### Reducing and clustering the data

The analysis can be considerably sped up, at little if any cost in accuracy, by reducing the dataset to only the top expressed genes (controlled by the `nfeatures` argument). Then, if the `clusters` argument isn't provided, the cells will be clustered using a fast clustering procedure on the PCA space (taken from the 'PCA' dimRed if present, otherwise generated internally). The aim, here, is to favour over-clustering so as to avoid collapsing pairs of celltypes whose doublets could be distinguishable. The internal clustering procedure can be accessed via the `fastcluster` function, and the cluster assignments are included in the output.

If the dataset is not expected to contain distinct subpopulations but rather continuous gradients, e.g. trajectories, then it might be advisable to employ a different approach and setting `clusters` to a positive integer (see below).

### Generating artificial doublets

In addition to a proportion of artificial doublets generated by combining random cells, a large fraction of the doublets are generated on all pairs of non-identical clusters (this can be performed manually using the `getArtificialDoublets` function). The rationale is that homotypic doublets are nearly impossible to distinguish without cellular barcodes, and therefore that creating that kind of doublets is a waste of computational resources. 

If the dataset is not expected to contain distinct subpopulations, but rather trajectories, a different approach can be used. A first option is to generate artificial doublets randomly, which can be done using the `propRandom=1` argument. A better approach is to use `trajectoryMode=TRUE`, and (unless you already have your clusters) setting the `clusters` argument of `scDblFinder` to a positive integer (depending on the number of cells and complexity, e.g. `k=12` for smaller datasets), which will split the cells using k-means clustering (where `k=clusters`). The average cluster profiles are then related, so that fewer doublets are created between very nearby points in the trajectory (where they would be wasted, such doublets being nearly undistinguishable from real cells).

### Examining the k-nearest neighbors (KNN) of each cell

A new PCA is performed on the combination of real and artificial cells, from which a KNN network is generated. Using this KNN, a number of parameters are gathered for each cell, such as the proportion of doublets (i.e. artificial doublets or known doublets provided through the `knownDoublets` argument, if given - these could be for instance inter-sample doublets flagged by SNPs) among the KNN, ratio of the distances to the nearest doublet and nearest non-doublet, etc. Several of this features are reported in the output with the 'scDblFinder.' prefix, e.g.:

* `distanceToNearest` : distance to the nearest cell (real or artificial)
* `nearestClass` : whether the nearest cell is a doublet or singlet
* `ratio` : the proportion of the KNN that are doublets. (If more than one value of `k` is given, the various ratios will be used during classification and will be reported)
* `weighted` : the proportion of the KNN that are doublets, weighted by their distance (useful for isolated cells)

### Training a classifier

Unless the `score` argument is set to 'weighted' or 'ratio' (in which case the aforementioned ratio is directly used as a doublet score), `scDblFinder` then uses gradient boosted trees trained on the KNN-derived properties along with a few additional features (e.g. library size, number of non-zero features, and an estimate of the difficultly of detecting artificial doublets in the cell's neighborhood, etc.) to distinguish doublets (either artificial or given) from other cells, and assigns a score on this basis. 
If the `use.cxds=TRUE`, the `cxds` score from the `r Biocpkg("scds","scds")` package will also be included among the predictors.

One problem of using a classifier for this task is that some of the real cells (the actual doublets) are mislabeled as singlet, so to speak. 
`scDblFinder` therefore iteratively retrains the classifier, each time excluding from the training the (real) cells called as doublets in the previous step (the number of steps being controlled by the `iter` parameter).

This score is available in the output, in the `scDblFinder.score` colData column, and can be interpreted as a probability. If the data is multi-sample, a single model is trained for all samples.

### Thresholding

Rather than thresholding on some arbitrary cutoff of the score, `scDblFinder` uses the expected number of doublets to establish a threshold. Unless it is manually given through the `dbr` argument, the expected doublet rate is first estimated using the empirical rule of thumb applicable to 10X data, namely roughly 1\% per 1000 cells captures (so with 5000 cells, (0.01\*5)\*5000 = 250 doublets, and the more cells you capture the higher the chance of creating a doublet). If samples were specified, and if the `dbr` is automatically calculated, thresholding is performed separately across samples.

Thresholding then tries to simultaneously minimize: 1) the classification error (in terms of the proportion of known doublets real cells below the threshold) and 2) the deviation from the expected number of doublets (as a ratio of the total number of expected doublets within the range determined by `dbr.sd`, and adjusted for homotypic doublets), giving an equal weight to each. This means that, if you have no idea about the doublet rate, setting `dbr.sd=1` will make the threshold depend entirely on the misclassification rate.

### Doublet origins and enrichments

Because we generate artificial doublets from clusters, it is most often possible to call the most likely origin (in terms of the combination of clusters) of a given putative real doublet. This information is provided through the `scDblFinder.mostLikelyOrigin` column of the output (and the `scDblFinder.originAmbiguous` column indicates whether this origin is ambiguous or rather clear). This, in turn, allows us to identify enrichment over expectation for specific kinds of doublets. Some statistics on each combination of clusters are saved in `metadata(sce)$scDblFinder.stats`, and the `plotDoubletMap` function can be used to visualize enrichments. Most enrichment or depletion is explained by the difficulty of identifying doublets of certain kinds, but departures from expectation might also point to technical or biological effects.

<br/><br/>

## Some important parameters

`scDblFinder` has a fair number of parameters governing the preprocessing, generation of doublets, classification, etc. (see `?scDblFinder`). Here we describe just a few of the most important ones.

### Expected proportion of doublets

The expected proportion of doublets has no impact on the density of artificial doublets in the neighborhood, but impacts the classifier's score and, especially, where the cutoff will be placed. It is specified through the `dbr` parameter and the `dbr.sd` parameter (the latter specifies a +/- range around `dbr` within which the deviation from `dbr` will be considered null). For 10x data, the more cells you capture the higher the chance of creating a doublet, and Chromium documentation indicates a doublet rate of roughly 1\% per 1000 cells captures (so with 5000 cells, (0.01\*5)\*5000 = 250 doublets), and the default expected doublet rate will be set to this value (with a default standard deviation of 0.015). Note however that different protocols may create considerably more doublets, and that this should be updated accordingly. If you have unsure about the doublet rate, you might consider increasing `dbr.sd`, so that it is estimated mostl/purely from the misclassification error.

### Clustering

Since doublets are created across clusters, it is important that subpopulations are not misrepresented as belonging to the same cluster. For this reason, we favor over-clustering at this stage, and if you provide your own clusters, the resolution should not be too coarse -- although an overly fine-grained resolution tends to reduce accuracy. `scDblFinder`'s default clustering method is implemented in the `fastcluster` function.

<br/><br/>

## Frequently-asked questions

### I'm getting way too many doublets called - what's going on?

Then you most likely have a wrong doublet rate. If you did not provide it (`dbr` argument), the doublet rate will be calculated automatically using expected doublet rates from 10x, meaning that the more cells captured, the higher the doublet rates. If you have reasons to think that this is not applicable to your data, set the `dbr` manually.

The most common cause for an unexpectedly large proportion of doublets is if you have a multi-sample dataset and did not split by samples. `scDblFinder` will think that the data is a single capture with loads of cells, and hence with a very high doublet rate. Splitting by sample should solve the issue.

The tresholding tries to minimize both the deviance from the expected number of doublets and the misclassification (i.e. of artificial doublets), meaning that the effective (i.e. final) doublet rate will differ from the given one. `scDblFinder` also considers false positive doublets to be less problematic than false negative. You can reduce to some degree the deviation from the input doublet rate by setting `dbr.sd=0`.

### The clusters don't make any sense!

If you ran `scDblFinder` on a multi-sample dataset and did not provide the cluster labels, then the labels are sample-specific (meaning that label '1' in one sample might have nothing to do with label '1' in another), and plotting them on a tSNE will look like they do not make sense. For this reason, when running multiple samples we recommend to first cluster all samples together (for example using `sce$cluster <- fastcluster(sce)`) and then provide the clusters to `scDblFinder`.

### Can I use this in combination with Seurat or other tools?

If the input SCE already contains a `logcounts` assay or a `reducedDim` slot named 'PCA', scDblFinder will use them for the clustering step. In addition, a clustering can be manually given using the `clusters` argument of `scDblFinder()`.
In this way, `r Githubpkg("satijalab.org/seurat")` clustering could for instance be used to create the artifical doublets (see `?Seurat::as.SingleCellExperiment.Seurat` for conversion to SCE).

### Can this be used with scATACseq data?

We have not yet thoroughly tested the paremeters in the context of scATACseq data, however preliminary results on a couple of datasets (applied on peak-level counts) suggest that it works decently there. However, since scATACseq data normally have more peaks with fewer reads, we suggest increasing the `nfeatures` used.