---
title: Scoring potential doublets from simulated densities
package: scDblFinder
author: 
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{computeDoubletDensity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# tl;dr

To demonstrate, we'll use one of the mammary gland datasets from the `r Biocpkg("scRNAseq")` package.
We will subset it down to a random set of 1000 cells for speed.

```{r}
library(scRNAseq)
sce <- BachMammaryData(samples="G_1")

set.seed(1001)
sce <- sce[,sample(ncol(sce), 1000)]
```

For the purposes of this demonstration, we'll perform an extremely expedited analysis.
One would usually take more care here and do some quality control,
create some diagnostic plots, etc., but we don't have the space for that.

```{r}
library(scuttle)
sce <- logNormCounts(sce)

library(scran)
dec <- modelGeneVar(sce)
hvgs <- getTopHVGs(dec, n=1000)

library(scater)
set.seed(1002)
sce <- runPCA(sce, ncomponents=10, subset_row=hvgs)
sce <- runTSNE(sce, dimred="PCA")
```

We run `computeDoubletDensity()` to obtain a doublet score for each cell based on the density of simulated doublets around it.
We log this to get some better dynamic range.

```{r}
set.seed(1003)
library(scDblFinder)
scores <- computeDoubletDensity(sce, subset.row=hvgs)
plotTSNE(sce, colour_by=I(log1p(scores)))
```

```{r, echo=FALSE}
# Sanity check that the plot has one cluster with much higher scores.
# If this fails, we probably need to pick a more demonstrative example.
library(bluster)
clusters <- clusterRows(reducedDim(sce, "PCA"), NNGraphParam())
by.clust <- split(scores, clusters)
med.scores <- sort(vapply(by.clust, median, 0), decreasing=TRUE)
stopifnot(med.scores[1] > med.scores[2] * 5)
```

# Algorithm overview {#overview}

We use a fairly simple approach in `doubletCells` that involves creating simulated doublets from the original data set:

1. Perform a PCA on the log-normalized expression for all cells in the dataset.
2. Randomly select two cells and add their count profile together.
Compute the log-normalized profile and project it into the PC space.
3. Repeat **2** to obtain $N_s$ simulated doublet cells.
4. For each cell, compute the local density of simulated doublets, scaled by the squared density of the original cells.
This is used as the doublet score.

# Size factor handling

## Normalization size factors

We allow specification of two sets of size factors for different purposes.
The first set is the normalization set: division of counts by these size factors yields expression values to be compared across cells.
This is necessary to compute log-normalized expression values for the PCA.

These size factors are usually computed from some method that assumes most genes are not DE.
We default to library size normalization though any arbitrary set of size factors can be used.
The size factor for each doublet is computed as the sum of size factors for the individual cells, based on the additivity of scaling biases.

## RNA content size factors

The second set is the RNA content set: division of counts by these size factors yields expression values that are proportional to absolute abundance across cells.
This affects the creation of simulated doublets by controlling the scaling of the count profiles for the individual cells.
These size factors would normally be estimated with spike-ins, but in their absence we default to using unity for all cells.

The use of unity values implies that the library size for each cell is a good proxy for total RNA content.
This is unlikely to be true: technical biases mean that the library size is an imprecise relative estimate of the content.
Saturation effects and composition biases also mean that the expected library size for each population is not an accurate estimate of content.
The imprecision will spread out the simulated doublets while the inaccuracy will result in a systematic shift from the location of true doublets.

Arguably, such problems exist for any doublet estimation method without spike-in information.
We can only hope that the inaccuracies have only minor effects on the creation of simulated cells.
Indeed, the first effect does mitigate the second to some extent by ensuring that some simulated doublets will occupy the neighbourhood of the true doublets.

## Interactions between them

These two sets of size factors play different roles so it is possible to specify both of them.
We use the following algorithm to accommodate non-unity values for the RNA content size factors:

1. The RNA content size factors are used to scale the counts first.
This ensures that RNA content has the desired effect in step **2** of Section \@ref(overview).
2. The normalization size factors are also divided by the content size factors.
This ensures that normalization has the correct effect, see below.
3. The rest of the algorithm proceeds as if the RNA content size factors were unity.
Addition of count profiles is done without further scaling, and normalized expression values are computed with the rescaled normalization size factors.

To understand the correctness of the rescaled normalization size factors, consider a non-DE gene with abundance $\lambda_g$.
The expected count in each cell is $\lambda_g s_i$ for scaling bias $s_i$ (i.e., normalization size factor).
The rescaled count is $\lambda_g s_i c_i^{-1}$ for some RNA content size factor $c_i$.
The rescaled normalization size factor is $s_i c_i^{-1}$, such that normalization yields $\lambda_g$ as desired.
This also holds for doublets where the scaling biases and size factors are additive.

# Doublet score calculations

## Theoretical basis

Consider a cell population with each subpopulation $x$ present in proportion $p_x$.
For a doublet rate $r$, we would expect to see self-doublets for subpopulation $x$ at a frequency of $rp_x^2$.
Inter-population doublets for subpopulations $x$ and $y$ should be observed at a frequency of $2rp_xp_y$.

We assume that $r$ is low such that the simulated doublets are generated at close-to-theoretical frequencies (i.e., negligible simulated doublets of real doublets).
To obtain a doublet score for each empirical cluster, we divide the number of simulated doublets mapped to each subpopulation by the squared cluster proportion. 
For a cluster corresponding to a real (i.e., non-doublet) subpopulation, this gives us a constant value equal to $r$.
For a doublet cluster, we should obtain $2(rp_xp_y)^{-1}$.
This should be large for $r \ll 1$ and $p_x, p_y < 1$, indicating that it is indeed composed of doublets.

We generalize this to each cell by:

1. Computing the density of simulated doublets neighbouring each cell.
This is a generalization for the number of simulated doublets mapped to each subpopulation, 
where the concept of a subpopulation is generalized to a region of the high-dimensional expression space.
2. Computing the density of original cells around each cell.
This is a generalization of the subpopulation size for this region of the expression space.
3. We divide the density of the simulated cells by the **squared** density of original cells to obtain a doublet score.
This can be used for relative comparisons between cells, with high scores indicating that a cell is from a doublet-like region of the expression space.

If `adjust=FALSE`, the squaring is omitted and the output score simply becomes the ratio of simulated to original densities.
This is mostly provided for debugging purposes.

## Density calculations

Previously, we computed the density by applying a tricube-weighted kernel to the distances to the neighbouring cells.
The bandwidth of the kernel is defined as the median distance to the 50th nearest neighbour across all cells.
The aim was to provide a threshold that adapts to the data and captures sufficient neighbours for stable density calculations.
We use a constant bandwidth to make it easier to compare densities between cells (no need for normalization, less imprecision).

Now, we simply count the proportion of cells in a hypersphere with radius set to the median distance to the 50th nearest neighbor.
This is deliberately insensitive to distances within the hypersphere.
In theory, the distance information should yield more accurate density estimates,
but in practice, it yields excessively high densities for doublet populations that have few cells but are very compact.

**Comment:** Many other doublet-based methods take a $k$-nearest neighbours approach to compute densities.
This is faster but is sensitive to the choice of $N_s$.
If $N_s$ is too large relative to the number of real cells, all of the $k$ nearest neighbours will be simulated, while if $N_s$ is too small, all of the nearest neighbors will be original cells.
Our density calculations separate considerations of speed and precision (dependent on $N_s$) from the expected value of the density (independent of $N_s$), which enables more predictable function behavior.

# Force matching

As mentioned above, there is a risk of mismatch between simulated and real doublets when RNA content is not considered.
This can be mitigated by forcing all simulated doublets to the closest neighbours in the original data set.
We identify the $k$ nearest original cells for each simulated doublet and we define new coordinates for the simulated doublet as the average profile across the $k$ neighbours.
(In practice, this is weighted by distance using a tricube kernel to avoid being skewed by distant cells in the $k$ nearest set.)

The force matching approach remaps simulated doublets to the closest cluster of original cells. 
This corrects for any systematic shift due to RNA content, assuming that the content-related shift is not severe enough to change the nearest neighbouring cluster.
The downside is that _all_ simulated doublets are mapped to their nearest original clusters.
This inflates the scores for all cells, possibility incorrectly if a cluster of simulated doublets is forced somewhere it should not be.

# Session information {-}

```{r}
sessionInfo()
```
