---
title: "scDblFinder"
author:
- name: Pierre-Luc Germain
  affiliation: University and ETH Zürich
package: scDblFinder
output:
  BiocStyle::html_document
abstract: |
  An introduction to the scDblFinder package, which identifies doublets in single-cell 
  RNAseq directly from counts using overclustering-based generation of artifical doublets.
vignette: |
  %\VignetteIndexEntry{scDblFinder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(BiocStyle)
```

# scDblFinder

scDblFinder identifies doublets in single-cell RNAseq by creating artificial doublets and looking at their
prevalence in the neighborhood of each cell, along with a few other covariates. The rough logic is very similar to 
other methods (e.g. `r Githubpkg("chris-mcginnis-ucsf/DoubletFinder")` or `r Biocpkg("scds")`), with a few twists 
that make it more efficient and provide extra features.

The approach described here is complementary to doublets identified via cell hashes and SNPs in multiplexed samples.  
The latter can identify doublets formed by cells of the same type from two samples, which are nearly undistinguishable
from real cells transcriptionally (and hence generally unidentifiable through the present package), but cannot 
identify doublets made by cells of the same sample. Moreover, when such inter-sample doublets are available, they
can be provided to `scDblFinder` through the `knownDoublets` argument to improve the identification of further doublets.


## Installation

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("scDblFinder")

# or, to get that latest developments:
BiocManager::install("plger/scDblFinder")
```

## Usage

Given an object `sce` of class `SingleCellExperiment` (which does not contain any empty drops, but hasn't been further filtered) :
```{r, message=FALSE, warning=FALSE}
library(scDblFinder)
# we create a dummy dataset
sce <- mockDoubletSCE()
```

We run `scDblFinder` simply with:

```{r}
sce <- scDblFinder(sce, verbose=FALSE)
```


This will add a number of columns to the colData of `sce` prefixed with 'scDblFinder', the most important of which are:

* `sce$scDblFinder.score` : the final doublet score
* `sce$scDblFinder.class` : the classification (doublet or singlet)

```{r}
table(sce$scDblFinder.class)
```


### Multiple samples

If you have multiple samples (understood as different cell captures), then it is
preferable to look for doublets separately for each sample (for multiplexed samples with cell hashes, 
this means for each batch). You can do this by simply providing a vector of the sample ids to the 
`samples` parameter of scDblFinder or, if these are stored in a column of `colData`, the name of the 
column. In this case, you might also consider multithreading it using the `BPPARAM` parameter. 
For example:

```{r, eval=FALSE}
library(BiocParallel)
sce <- scDblFinder(sce, samples="sample_id", BPPARAM=MulticoreParam(3))
table(sce$scDblFinder.class)
```

<br/><br/>

## Description of the method

Wrapped in the `scDblFinder` function are the following steps:

### Splitting captures

Doublets can only arise within a given sample or capture, and for this reason are better sought independently for each sample, which also speeds up the analysis. If the `samples` argument is given, `scDblFinder` will use it to split the cells into samples/captures, and process each of them (in parallel if the `BPPARAM` argument is given). If your samples are multiplexed, i.e. the different samples are mixed in different batches, then the batches should be what you provide to this argument.

### Reducing and clustering the data

The analysis can be considerably sped up, at little if any cost in accuracy, by reducing the dataset to only the top expressed genes (controlled by the `nfeatures` argument). Then, if the `clusters` argument isn't provided, the cells will be clustered using a fast clustering procedure on the PCA space (taken from the 'PCA' dimRed if present, otherwise generated). The method is controlled by the `clust.method` argument. The aim, here, is to favour over-clustering so as to avoid collapsing pairs of celltypes whose doublets could be distinguishable. The clusters are included in the output.

### Generating artificial doublets

Artificial doublets are then generated on all pairs of non-identical clusters (homotypic doublets being nearly impossible to distinguish without cellular barcodes). This can be performed manually using the `getArtificialDoublets` function.

### Examining the k-nearest neighbors (KNN) of each cell

A new PCA is performed on the combination of real and artificial cells, from which a KNN network is generated. Using this KNN, a number of parameters are gathered for each cell, such as the proportion of doublets (i.e. artificial doublets or known doublets provided through the `knownDoublets` argument, if given - these could be for instance inter-sample doublets flagged by SNPs) among the KNN, ratio of the distances to the nearest doublet and nearest non-doublet, etc. Several of this features are reported in the output with the 'scDblFinder.' prefix:

* `distanceToNearest` : distance to the nearest cell
* `nearestClass` : whether the nearest cell is a doublet or singlet
* `ratio` : the proportion of the KNN that are doublets
* `weighted` : the proportion of the KNN that are doublets, weighted by their distance (particularly useful for isolated cells)

### Training a classifier

Unless the `score` argument is set to 'weighted' or 'ratio' (in which case the aforementioned ratio is directly used as a doublet score), `scDblFinder` then uses gradient boosted trees trained on the KNN-derived properties along with a few additional features (e.g. library size, number of non-zero features, and an estimate of the difficultly of detecting artificial doublets in the cell's neighborhood) to distinguish doublets (either artificial or given) from other cells, and assigns a score on this basis. This score is available in the output as either `scDblFinder.score` or `scDblFinder.score.global` (see below).

### Local calibration and thresholding

Rather than thresholding on some arbitrary cutoff of the score, `scDblFinder` uses the expected number of doublets to establish a threshold. Unless it is manually given through the `dbr` argument, the expected doublet rate is first estimated using the empirical rule of thumb applicable to 10X data, namely roughly 1\% per 1000 cells captures (so with 5000 cells, (0.01\*5)\*5000 = 250 doublets, and the more cells you capture the higher the chance of creating a doublet). This doublet rate is then adjusted so that the expected amount of homotypic doublets, based on the relative cluster prevalences, is removed.

Thresholding then tries to simultaneously minimize: 1) the classification error (in terms of the proportion of known doublets below the threshold) and 2) the deviation from the expected number of doublets (as a ratio of the total number of expected doublets), giving an equal weight to each (although in practice the deviation from the expected number of doublets will be the biggest driver of the threshold). If `score` is either to 'xgb' (default), 'weighted' or 'ratio', then thresholding is performed directly on the said score. If `score='xgb.local.optim'`, then a local calibration of the score is performed.

#### Local calibration (optional)

Doublets that are located in an otherwise sparse region of the space are easier to identify (and hence will have a higher score) that those located in a dense region. In other words, the aforementioned doublet scores are biased against rarer cell states. To correct for this, it is possible to perform threshold optimization separately for different groups of cells, depending on the type of artificial doublets (i.e. their originating clusters) closest to them. In this context, the expected number of doublets is calculated for each combination of clusters.

This procedure can be problematic whenever the doublets are not random (e.g. when two cell types preferentially stick together). For this reason, instead of using the local threshold directly, `scDblFinder` moderates it by establishing a relationship between the local threshold and the difficultly in indentifying doublets, and pulling the threshold towards this expectation.

The cells' scores are then transformed so that the different local thresholds are equalized, by substracting the logit-threshold to the logit-score and transforming back to a 0-1 score. Finally, the global threshold is calculated as described above. When the local calibration is enabled, the calibrated score is available in `scDblFinder.score`, and the original one in `scDblFinder.score.global`.

### Doublet origins and enrichments

Because the artificial doublets are generated from clusters, it is most often possible to call the most likely origin (in terms of the combination of clusters) of a given putative real doublet. This information is provided through the `scDblFinder.mostLikelyOrigin` column of the output (and the `scDblFinder.originAmbiguous` column indicates whether this origin is ambiguous or rather clear). This, in turn, allows us to identify enrichment over expectation for specific kinds of doublets. Some statistics on each combination of clusters are saved in `metadata(sce)$scDblFinder.stats`, and the `plotDoubletMap` function can be used to visualize enrichments. Most enrichment or depletion is explained by the difficulty of identifying doublets of certain kinds, but departures from expectation might also point to technical or biological effects.

<br/><br/>

## Important parameters

The important sets of parameters in `scDblFinder` refer respectively to the expected proportion of doublets, to the clustering, and to the number of artificial doublets used.

### Expected proportion of doublets

The expected proportion of doublets has no impact on the score (the `ratio` above), but a very strong impact on where the threshold will be placed. It is specified through the `dbr` parameter and the `dbr.sd` parameter (the latter specifies the standard deviation of `dbr`, i.e. the uncertainty in the expected doublet rate). For 10x data, the more cells you capture the higher the chance of creating a doublet, and Chromium documentation indicates a doublet rate of roughly 1\% per 1000 cells captures (so with 5000 cells, (0.01\*5)\*5000 = 250 doublets), and the default expected doublet rate will be set to this value (with a default standard deviation of 0.015). Note however that different protocols may create considerably more doublets, and that this should be updated accordingly.

### Clustering

Since doublets are created across clusters, it is important that subpopulations are not misrepresented as belonging to the same cluster. For this reason, we favor over-clustering at this stage. scDblFinder's default clustering method is implemented in the `fastcluster` function; alternatives include the `overcluster` function (specifying minimum and maximum cluster sizes) and manually providing cluster labels.

### Number of artificial doublets

`scDblFinder` itself determines a reasonable number of artificial doublets to create on the basis of the size of the population and the number of clusters, but increasing this number can only increase the accuracy.

<br/><br/>

## Combination with other tools

If the input SCE already contains a `logcounts` assay or a `reducedDim` slot named 'PCA', scDblFinder will use them for the clustering step. In addition, a clustering can be manually given using the `clusters` argument of `scDblFinder()`. In this way, `r Githubpkg("satijalab.org/seurat")` clustering could for instance be used (in which case we suggest to increase the `resolution` parameter) to create the artifical doublets (see `?Seurat::as.SingleCellExperiment.Seurat` for conversion to SCE).
