---
title: "scDblFinder"
author:
- name: Pierre-Luc Germain
  affiliation: University and ETH Zürich
package: scDblFinder
output:
  BiocStyle::html_document
abstract: |
  An introduction to the scDblFinder package, which identifies doublets in single-cell 
  RNAseq directly from counts using overclustering-based generation of artifical doublets.
vignette: |
  %\VignetteIndexEntry{scDblFinder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(BiocStyle)
```

# scDblFinder

## Introduction

scDblFinder identifies doublets in single-cell RNAseq directly by creating artificial doublets and looking at their
prevalence in the neighborhood of each cell. The rough logic is very similar to
`r Githubpkg("chris-mcginnis-ucsf/DoubletFinder")`, but it is simpler and more efficient. In a
nutshell, instead of creating doublets from random pairs of cells, scDblFinder first overclusters the cells and
create cross-cluster doublets. It also uses meta-cells from each cluster to create triplets. This strategy avoids
creating homotypic doublets and enables the detection of most heterotypic doublets with much fewer artificial doublets.
We also rely on the expected proportion of doublets to threshold the scores, we include a variability in the estimate
of the doublet proportion (`dbr.sd`), and use the error rate of the real/artificial predicition in conjunction with
the deviation in global doublet rate to set the threshold.

The approach described here is complementary to doublets identified via cell hashes and SNPs in multiplexed samples.  
The latter can identify doublets formed by cells of the same type from two samples, which are nearly undistinguishable
from real cells transcriptionally (and hence unidentifiable through the present package), but cannot identify doublets
made by cells of the same sample.

## Installation

scDblFinder was developed under R 3.6. Install with:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("scDblFinder")

# or, to get that latest developments:
BiocManager::install("plger/scDblFinder")
```

## Usage

Given an object `sce` of class `SingleCellExperiment` (which does not contain any empty drops, but hasn't been further filtered) :
```{r}
# we create a dummy dataset
m <- t(sapply( seq(from=0, to=5, length.out=500), 
                FUN=function(x) rpois(1000,x) ) )
# we add a few random artificial doublets:
m <- cbind(m, m[,1]+m[,2], m[,20]+m[,30])
sce <- SingleCellExperiment( list(counts=m) )
sce <- scDblFinder(sce, verbose=FALSE)
```

We run `scDblFinder` simply with:

```{r}
sce <- scDblFinder(sce, verbose=FALSE)
```


This will add a number of columns to the colData of `sce` prefixed with 'scDblFinder', the most important of which are:

* `sce$scDblFinder.score` :  the final doublet score
* `sce$scDblFinder.class` : the classification (doublet or singlet)

```{r}
table(sce$scDblFinder.class)
```


### Multiple samples

If you have multiple samples (understood as different cell captures), then it is
preferable to look for doublets separately for each sample (for multiplexed samples with cell hashes, 
this means for each batch). You can do this by simply providing a vector of the sample ids to the 
`samples` parameter of scDblFinder or, if these are stored in a column of `colData`, the name of the 
column. In this case, you might also consider multithreading it using the `BPPARAM` parameter. 
For example:

```{r, eval=FALSE}
library(BiocParallel)
sce <- scDblFinder(sce, samples="sample_id", BPPARAM=MulticoreParam(3))
table(sce$scDblFinder.class)
```


### Parameters

The important sets of parameters in `scDblFinder` refer respectively to the expected proportion of doublets, to the clustering, and to the number of artificial doublets used.

#### Expected proportion of doublets

The expected proportion of doublets has no impact on the score (the `ratio` above), but a very strong impact on where the threshold will be placed. It is specified through the `dbr` parameter and the `dbr.sd` parameter (the latter specifies the standard deviation of `dbr`, i.e. the uncertainty in the expected doublet rate). For 10x data, the more cells you capture the higher the chance of creating a doublet, and Chromium documentation indicates a doublet rate of roughly 1\% per 1000 cells captures (so with 5000 cells, (0.01\*5)\*5000 = 250 doublets), and the default expected doublet rate will be set to this value (with a default standard deviation of 0.015). Note however that different protocols may create considerably more doublets, and that this should be updated accordingly.

#### Clustering

Since doublets are created across clusters, it is important that subpopulations are not misrepresented as belonging to the same cluster. For this reason, we favor over-clustering at this stage. scDblFinder's default clustering method is implemented in the `fastcluster` function; alternatives include the `overcluster` function (specifying minimum and maximum cluster sizes) and manually providing cluster labels.

#### Number of artificial doublets

`scDblFinder` itself determines a reasonable number of artificial doublets to create on the basis of the size of the population and the number of clusters, but increasing this number can only increase the accuracy.

<br/><br/>

## Combination with other tools

If the input SCE already contains a `logcounts` and/or `reducedDim` slot named 'PCA', scDblFinder will used them for the clustering step. In addition, a clustering can be manually given using the `clusters` argument of `scDblFinder()`. In this way, `r Githubpkg("satijalab.org/seurat")` clustering could for instance be used (in which case we suggest to increase the `resolution` parameter) to create the artifical doublets (see `?Seurat::as.SingleCellExperiment.Seurat` for conversion to SCE).
